---
title: "Просторовий аналіз умов праці в Україні"
subtitle: "Статистика праці за областями (2015-2019 рр.))"
author: "Mykola Utkin"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: show
    theme: cosmo
    fig-width: 12
    fig-height: 10
---

## Introduction

This analysis explores the spatial distribution of working conditions across Ukrainian oblasts using official labor statistics from 2015-2019. We examine the share of employees working in harmful conditions and visualize regional disparities through choropleth maps.

**Research Objectives:**

1. Map the percentage of workers in harmful working conditions by oblast
2. Identify regional patterns and disparities in workplace safety
3. Analyze temporal changes between 2015 and 2019
4. Visualize industrial concentration and its relationship to working conditions

## Setup and Data Loading

```{r setup, message=FALSE, warning=FALSE}
# Load required packages
if (!require("pacman")) install.packages("pacman")
pacman::p_load(
  sf,
  tidyverse,
  giscoR,
  rnaturalearth,
  scales,
  viridis,
  patchwork,
  knitr,
  here
)

# Set options
options(scipen = 999)  # Disable scientific notation
```

## Step 1: Load Spatial Data

```{r load-spatial}
# Load Ukraine oblasts shapefile
ukraine_oblasts <- read_sf("data/geoBoundaries-UKR-ADM1.shp")

# Display structure
glimpse(ukraine_oblasts)

# Check unique oblast names
cat("Number of regions:", nrow(ukraine_oblasts), "\n")
cat("\nOblast names:\n")
print(sort(unique(ukraine_oblasts$shapeName)))
```

### Standardize Oblast Names

```{r standardize-names}
# Create standardized names matching the tabular data
ukraine_oblasts <- ukraine_oblasts |>
  mutate(
    oblast_standard = case_when(
      shapeName == "Volyn Oblast" ~ "Волинська",
      shapeName == "Rivne Oblast" ~ "Рівненська",
      shapeName == "Zhytomyr Oblast" ~ "Житомирська",
      shapeName == "Kyiv Oblast" ~ "Київська",
      shapeName == "Chernihiv Oblast" ~ "Чернігівська",
      shapeName == "Sumy Oblast" ~ "Сумська",
      shapeName == "Kharkiv Oblast" ~ "Харківська",
      shapeName == "Luhansk Oblast" ~ "Луганська",
      shapeName == "Donetsk Oblast" ~ "Донецька",
      shapeName == "Zaporizhia Oblast" ~ "Запорізька",
      shapeName == "Kherson Oblast" ~ "Херсонська",
      shapeName == "Autonomous Republic of Crimea" ~ "АР Крим",
      shapeName == "Sevastopol" ~ "Севастополь",
      shapeName == "Mykolaiv Oblast" ~ "Миколаївська",
      shapeName == "Odessa Oblast" ~ "Одеська",
      shapeName == "Vinnytsia Oblast" ~ "Вінницька",
      shapeName == "Cherkasy Oblast" ~ "Черкаська",
      shapeName == "Poltava Oblast" ~ "Полтавська",
      shapeName == "Dnipropetrovsk Oblast" ~ "Дніпропетровська",
      shapeName == "Kirovohrad Oblast" ~ "Кіровоградська",
      shapeName == "Khmelnytskyi Oblast" ~ "Хмельницька",
      shapeName == "Ternopil Oblast" ~ "Тернопільська",
      shapeName == "Lviv Oblast" ~ "Львівська",
      shapeName == "Ivano-Frankivsk Oblast" ~ "Івано-Франківська",
      shapeName == "Zakarpattia Oblast" ~ "Закарпатська",
      shapeName == "Chernivtsi Oblast" ~ "Чернівецька",
      shapeName == "Kyiv" ~ "Київ",
      TRUE ~ shapeName
    )
  )

# Preview map
ggplot(ukraine_oblasts) +
  geom_sf(fill = "lightblue", color = "white", size = 0.5) +
  theme_void() +
  labs(title = "Ukraine: Administrative Divisions (Oblasts)")
```

## Step 2: Load and Clean Tabular Data

```{r load-tabular}
#| include: FALSE
# Load working conditions data
work_data <- read_csv("data/Work_conditions.csv", show_col_types = FALSE)

# Display structure
glimpse(work_data)
cat("\nFirst few rows:\n")
head(work_data) |> kable()
```

### Data Cleaning and Transformation